---
title: 'EDS 223: assignment 2'
author: "{Sujan}"
output: html_document
date: "2023-10-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Present-day environmental justice may reflect legacies of injustice in the past. The United States has a long history of racial segregation which is still visible. During the 1930's the Home Owners' Loan Corporation (HOLC), as part of the New Deal, rated neighborhoods based on their perceived safety for real estate investment. Their ranking system, (A (green), B (blue), C (yellow), D (red)) was then used to block access to loans for home ownership. Colloquially known as "redlining", this practice has had widely-documented consequences not only for community wealth, but also health.^[Gee, G. C. (2008). A multilevel analysis of the relationship between institutional and individual racial discrimination and health status. American journal of public health, 98(Supplement_1), S48-S56.] Redlined neighborhoods have less greenery^[Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A. (2021). Redlines and greenspace: the relationship between historical redlining and 2010 greenspace across the United States. Environmental health perspectives, 129(1), 017006.] and are hotter than other neighborhoods.^[Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects of historical housing policies on resident exposure to intra-urban heat: a study of 108 US urban areas. Climate, 8(1), 12.] 

Check out coverage by the [New York Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the environments communities are exposed to, it has also shaped our observations of biodiversity.^[Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical redlining is associated with increasing geographical disparities in bird biodiversity sampling in the United States. Nature Human Behaviour, 1-9.] Community or citizen science, whereby individuals share observations of species, is generating an enormous volume of data. Ellis-Soto and co-authors found that redlined neighborhoods remain the most undersampled areas across 195 US cities. This gap is highly concerning, because conservation decisions are made based on these data.

Check out coverage by [EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).


### Data
#### EJScreen
We will be working with data from the United States Environmental Protection Agency's EJScreen: Environmental Justice Screening and Mapping Tool.

According to the [US EPA website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

>This screening tool and data may be of interest to community residents or other stakeholders as they search for environmental or demographic information. It can also support a wide range of research and policy goals. The public has used EJScreen in many different locations and in many different ways.
>
>EPA is sharing EJScreen with the public:  
> - to be more transparent about how we consider environmental justice in our work,  
> - to assist our stakeholders in making informed decisions about pursuing environmental justice and,   
> - to create a common starting point between the agency and the public when looking at issues related to environmental justice. 

EJScreen provides on environmental and demographic information for the US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and [block group](https://en.wikipedia.org/wiki/Census_block_group) levels. You will be working with block group data that has been downloaded from the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). To understand the associated data columns, you will need to explore the Technical Documentation and column description spreadsheet available in the data folder. I also encourage you to explore the [limitations and caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen) of the data.

#### Mapping Inequality
A team of researchers, led by the [Digital Scholarship Lab](https://dsl.richmond.edu/) at the University of Richmond have digitized maps and information from the HOLC as part of the [Mapping Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58) project. 

We will be working with maps of HOLC grade designations for Los Angeles. Information on the data can be found [here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).^[Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan Connolly, et al., “Mapping Inequality,” American Panorama, ed. Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023, https://dsl.richmond.edu/panorama/redlining/] 

#### Biodiversity observations
The [Global Biodiversity Information Facility](gbif.org) is the largest aggregator of biodiversity observations in the world. Observations typically include a location and date that a species was observed.

We will be working observations of birds from 2021 onward.


## Assignment

### Investigate the legacy of redlining in current environmental (in)justice
Load relevant packages.
```{r }
library("sf")
library("tidyverse")
library("spDataLarge")
library("tmap")
library("rmapshaper")
```

Read in EJScreen data and filter to Los Angeles County
(5 points)
```{r }
##read the environmental justic screening data for Los Angeles county
ejscreen <- st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb/")

# filter for los angeles county
los_angeles <- ejscreen %>%
               filter(CNTY_NAME == "Los Angeles County")
```

Make a map of wastewater discharge by census block groups. Indicate which census block groups are above the 95th percentile of national values for wastewater discharge by adding a centroid.
(10 points)
```{r }
#wastewater discharge is mentioned as PWDIS
ggplot(los_angeles) +
  geom_sf(aes(color = PWDIS))

##create centroid, further filter them to have only 0.95 quantile
la_centroids <- st_centroid(los_angeles)
la_centroid_with_95percentile <- quantile(la_centroids$PWDIS, 0.95, na.rm = TRUE)

##fitler where PWDIS is greater than la_centroids_with_95_percentile
la_centroids_filtered <- la_centroids %>% 
                         filter(PWDIS > la_centroid_with_95percentile)

##add centroid 
ggplot(los_angeles) +
  geom_sf(aes(fill = PWDIS))+
  geom_sf(data = la_centroids_filtered, color = "red", size = 0.5)+
  theme_bw()
  
```

Find the percent of census block groups that have:  
- less than 5% of the population is considered low income
(5 points)
```{r }
##filter for rows where LOWINCPCT is less than 5%
less_than_5percent <- los_angeles %>% filter(LOWINCPCT < 0.05)

##Find the proportion using dplyr

proportion_calucalated <- los_angeles %>% 
                          summarize(proportion = nrow(less_than_5percent)/nrow(los_angeles)*100) %>% 
                          pull(proportion)

##diplay the answer
proportion_calucalated

```

Find the percent of census block groups that are:    
- above the 80th percentile for Particulate Matter 2.5 AND  
- above the 80th percentile for Superfund proximity
(10 points)
```{r }
##first create the new column that calculates percentiles for values
PM2.5_greater_than_eighty <- los_angeles %>%
                             filter(PM25 > quantile(los_angeles$PM25, 0.8, na.rm = TRUE))

proportion_pm25 <- los_angeles %>% 
              summarize(proportion = nrow(PM2.5_greater_than_eighty)/nrow(los_angeles)*100) %>% 
              pull(proportion)

proportion_pm25

##for the superfund proximity
superfund_greater_than_eighty <- los_angeles %>%
                                 filter(PNPL > quantile(los_angeles$PNPL, 0.8, na.rm = TRUE))

proportion_superfund <- los_angeles %>% 
              summarize(proportion = nrow(superfund_greater_than_eighty)/nrow(los_angeles)*100) %>% 
              pull(proportion)

proportion_superfund


```

Import redlining information for Los Angeles.
```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/downloads/geojson/CALosAngeles1939.geojson") %>%
  st_make_valid()

```

Make a map of historical redlining boundaries, colored by HOLC grade.
(5 points)
```{r }
redlining <- ggplot(LA_redlining)+
             geom_sf(aes(fill = holc_grade))+
             theme_bw()+
             labs(fill = "Redlining area Environmental Justice Screening")

redlining
```

Find the number of census block groups that fall within areas with HOLC grades
hint: make sure the CRS match
(15 points)
```{r eval=FALSE}
##subsetting based on within values

##transform crs one to another to match
los_angeles_transformed <- st_transform(los_angeles, crs = st_crs(LA_redlining))

##check if they are valid geometry objects
LA_redlining <- st_make_valid(LA_redlining)
los_angeles_transformed <- st_make_valid(los_angeles_transformed)

#####based on this method, there are only 887 regions within redlining
container <- st_within(los_angeles_transformed, LA_redlining)
logical_values_containing_redlabling <- lengths(container) > 0


###make the dataframe
dataframe_with_redlabeled_region <- los_angeles_transformed[logical_values_containing_redlabling, ]
nrow(dataframe_with_redlabeled_region)

# #Using intersection method
# 
# ##  use intersect method to select only those areas where redlining was counted
# la_subset <- los_angeles_transformed[LA_redlining, ]
# 
# tm_shape(la_subset) +
#   tm_fill()
#   tm_shape(world_random) +
#   tm_fill(fill = "red")
```

Summarize current conditions based on EJScreen data within historical redlining categories using the mean of the following variables:  
-% low income.  
- percentile for particulate Matter 2.5.  
- percentile for low life expectancy.  
- percentile for air toxics cancer risk
(20 points)
```{r }
##Inner join the redline data with census block subset 
random_joined  <- st_join(dataframe_with_redlabeled_region, LA_redlining) %>% 
                  group_by(holc_grade) %>% 
                  summarize(mean_of_low_income = mean(LOWINCPCT),
                            mean_of_pm2.5_percentile = mean(P_PM25),
                            mean_of_low_life_Expectancy = mean(P_LIFEEXPPCT, na.rm = TRUE),
                            mean_of_p_cancer = mean(P_CANCER))
                         
```

Please write 2-3 sentences reflecting on these results.
(5 points)

Based on the data, there seems to have strong environmental racism in the LA area. The worse graded area like Grade B area and Grade C area
have the highest percentile for low life expectancy, cancer, and air pollution. In contrast, A graded area has comparatively low values for
all of these parameters, suggesting that historically redlined areas are not good residential areas even in present context.


### Investigate the legacy of redlining in biodiversity observations

For bird observations from 2022 that fall within neighborhoods with HOLC grads, find the percent of observations within each redlining categories and plot results.
hint: make sure that the bird observations have the same CRS as redlining data.
(20 points)
````{r }
bird <- st_read("data/gbif-birds-LA/gbif-birds-LA.shp") %>% 
        filter(year >= 2022)

##change the CRS, even without chekcing
bird_crs_transformed <- st_transform(bird, crs = st_crs(dataframe_with_redlabeled_region))

## join the data


```

Please write 2-3 sentences reflecting on these results. Are they surprising? If so, why? What are some potential implications?
(5 points)