---
Project: 'Environmental Racism and Biodiversity integrity'
author: "{Sujan}"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930's the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York
Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.[^4] Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Check out coverage by
[EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

### Data

I will be working with data from the United States Environmental
Protection Agency's EJScreen: Environmental Justice Screening and
Mapping Tool.

According to the [US EPA
website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents
> or other stakeholders as they search for environmental or demographic
> information. It can also support a wide range of research and policy
> goals. The public has used EJScreen in many different locations and in
> many different ways.
>
> EPA is sharing EJScreen with the public:\
> - to be more transparent about how we consider environmental justice
> in our work,\
> - to assist our stakeholders in making informed decisions about
> pursuing environmental justice and,\
> - to create a common starting point between the agency and the public
> when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels.
You will be working with block group data that has been downloaded from
the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). To
understand the associated data columns, you will need to explore the
Technical Documentation and column description spreadsheet available in
the data folder. I also encourage you to explore the [limitations and
caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen)
of the data.

#### Mapping Inequality

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

I will be working with maps of HOLC grade designations for Los Angeles.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

#### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed.

I will be working observations of birds from 2021 onward.

### Investigate the legacy of redlining in current environmental (in)justice

Load relevant packages.

```{r }
library(sf)
library(raster)
library(terra)
library(spData)
library(spDataLarge)
library(tidyverse)
library(ggspatial)
library(patchwork)
library(ggmap)
library(tmap)
```

# Part 1: Data Exploration

The part 1 explores the counties with the highest waste-water discharge,
percentage low-income people in each county, and low superfund
proximity. It also includes combined statistics for areas with high air
pollution, coupled with low superfund proximity. This step is more of a
data exploration to get general overview.

Read in EJScreen data and filter to Los Angeles County

```{r }
##read the environmental justic screening data for Los Angeles county
ejscreen <- st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb/")

# filter for los angeles county
los_angeles <- ejscreen %>%
               filter(CNTY_NAME == "Los Angeles County")
```

Make a map of wastewater discharge by census block groups. Indicate
which census block groups are above the 95th percentile of national
values for wastewater discharge by adding a centroid.

```{r fig.width= 5, fig.height=10 }
##create centroid, further filter them to have only 0.95 quantile
la_centroids <- st_centroid(los_angeles)
la_centroid_with_95percentile <- quantile(la_centroids$PWDIS, 0.95, na.rm = TRUE)

##fitler where PWDIS is greater than la_centroids_with_95_percentile
la_centroids_filtered <- la_centroids %>% 
                         filter(PWDIS > la_centroid_with_95percentile)
```

```{r fig.width= 5, fig.height=10 }
##add centroid 
ggplot(los_angeles) +
  geom_sf(aes(fill = PWDIS))+
  geom_sf(data = la_centroids_filtered, color = "red", size = 0.5)+
  theme_bw()+
  annotation_scale(plot_unit = "m")+
  annotation_north_arrow( location = "br", 
                          height = unit(1, "cm"))+
  ggtitle("Wasterwater Discharge for Los Angeles county")+
  xlab("Longitude")+
  ylab("Latitude")
```

Percent of census block groups that have less than 5% of the population is considered low income.

```{r }
##filter for rows where LOWINCPCT is less than 5%
less_than_5percent <- los_angeles %>% filter(LOWINCPCT < 0.05)

##Find the proportion using dplyr
proportion_calucalated <- los_angeles %>% 
                          summarize(proportion = nrow(less_than_5percent)/nrow(los_angeles)*100) %>% 
                          pull(proportion)
```

Percent of census block groups that are above the 80th percentile for Particulate Matter 2.5 AND above the
80th percentile for Superfund proximity.

```{r }
##using percentile coumns for air pollution and superfund proximity
pm_superfund <- los_angeles %>% 
                filter(P_PM25 > 80 & P_PNPL > 80)

proportion <- los_angeles %>% 
              summarize(prop = nrow(pm_superfund)/nrow(los_angeles)*100) %>% 
              pull(prop)
```

Import redlining information for Los Angeles county. This dataset is
more important as it categorizes the counties by HOLC grade, which is
important later when assessing biodiversiyt. In the meantime, this
dataset can be used to explore counties that fall in this grading.

```{r }
LA_redlining <- st_read(
  "https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json") %>%
  st_make_valid() %>% 
  filter(!is.na(grade))
```

```{r }
redlining <- ggplot(LA_redlining)+
             geom_sf(aes(fill = grade))+
             theme_bw()+
             labs(fill = "Redlining area Environmental Justice Screening")+
  annotation_scale(plot_unit = "km")+
  annotation_north_arrow( location = "tr", 
                          width = unit(0.5, "cm"))+
  ggtitle("Redlining areas in LA county")+
  xlab("Longitude")+
  ylab("Latitude")

redlining
```

finding number of census block groups that fall within areas with HOLC grades.
```{r}
##transform crs one to another to match
los_angeles_transformed <- st_transform(los_angeles, crs = st_crs(LA_redlining))

##check if they are valid geometry objects
LA_redlining <- st_make_valid(LA_redlining)
los_angeles_transformed <- st_make_valid(los_angeles_transformed)

#####based on this method, there are only 887 regions within redlining
container <- st_within(los_angeles_transformed, LA_redlining)
logical_values_containing_redlabling <- lengths(container) > 0

###make the dataframe
dataframe_with_redlabeled_region <- los_angeles_transformed[logical_values_containing_redlabling, ]
nrow(dataframe_with_redlabeled_region)
```

Summarize current conditions based on EJScreen data within historical
redlining categories using the mean of air pollution, life expectancy and air toxic cancer risk.

```{r }
##Inner join the redline data with census block subset 
random_joined  <- st_join(dataframe_with_redlabeled_region, LA_redlining) %>% 
                  group_by(grade) %>% 
                  summarize(mean_of_low_income = mean(LOWINCPCT),
                            mean_of_pm2.5_percentile = mean(P_PM25),
                            mean_of_low_life_Expectancy = mean(P_LIFEEXPPCT, na.rm =TRUE),
                            mean_of_p_cancer = mean(P_CANCER))
random_joined
```

Results from analysis:

Based on the data, there seems to have strong environmental racism in
the LA area. The worse graded area like Grade B area and Grade C area
have the highest percentile for low life expectancy, cancer, and air
pollution. In contrast, A graded area has comparatively low values for
all of these parameters, suggesting that historically redlined areas are
not good residential areas even in present context.

## Part 2: Data analysis and Decision Making

For bird observations from 2022 that fall within neighborhoods with HOLC
grads, find the percent of observations within each redlining categories
and plot results.

```{r}
bird_in_holc_grade <- st_read("data/gbif-birds-LA/gbif-birds-LA.shp") %>%
  filter(year >= 2022)
```

Prepare the dataset for the visualizaion:

```{r}
##read the shape file for the birds diveristy dataset
bird <- st_read("data/gbif-birds-LA/gbif-birds-LA.shp") %>% 
        filter(year >= 2022)

##change the CRS, even without chekcing
bird_crs_transformed <- st_transform(bird_in_holc_grade, crs = st_crs(LA_redlining))

##subset only redlined areas from bird dataset
bird_subset <- bird_crs_transformed[LA_redlining, ]

## join the data with redlining using st_join, default is inner join
bird_redlined <- st_join(LA_redlining, bird_subset)

result <- bird_redlined %>% group_by(grade) %>% 
          mutate(percent_of_birds = as.character(round(n()/nrow(bird_redlined)*100), 2))
```

```{r}
########aggregate the data and plot the summarized output 
tm_shape(LA_redlining)+
    tm_fill(fill = "grade", palette = c("green", "blue", "yellow", 'red'), alpha = 1)+
    tm_shape(result)+
    tm_dots(col = 'percent_of_birds', palette=c('15' ='red', '19'='yellow', '29'='blue', '36' ='green'))+
    tm_scale_bar(location = c("bottomright"), position = c("left", "bottom"))+
    tm_compass(type = "arrow", position = c("right", "top"))+
  tm_layout(title = "Redlinining areas and bird diversity index average")

```

Based on the graph, Most A graded areas(green areas in the graph) showed
low bird diversity (denoted by red dots) than B, C and D graded areas.
The potential implication of this finding is that low graded areas offer
unique habitat to bird species, thus policy makers should incorporate
these type of findings so that biodiversity protection is ensured.
Moreover, this finding challenges the belief that low graded areas have
low species diversity.
